Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2021-11-12 docs/00-manuscripts/mdpi/manuscript.tex

% Explain what is new in this study and link to methodology
Despite their popularity in environmental modeling, there are no studies so far which used ML algorithms in combination with remote sensing data to analyze defoliation at the tree level.
This study aims to close this gap by analyzing tree defoliation in northern Spain using airborne hyperspectral data.
The methodology of this study uses ML methods in combination with feature selection and hyperparameter tuning.
In addition, feature importance was analyzed.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high dimensionality in environmental modeling \cite{trunk1979, xu2016}.

@@

Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata D. Don}) plantations in northern Spain.
The trees in the plots suffer from infections of pathogens such as \textit{Diplodia sapinea (Fr.) Fuckel}, \textit{Fusarium circinatum Nirenberg \& O'Donnell}, \textit{Armillaria mellea (Vahl) P. Kumm}, \textit{Heterobasidion annosum (Fr.) Bref}, \textit{Lecanosticta acicola (Thüm) Syd.} and \textit{Dothisthroma septosporum (Dorogin) M. Morelet} causing (among others) needle blight, pitch canker and root diseases \cite{mesanza2016, iturritxa2017}.
The first two fungi are mainly responsible for the foliation loss of the trees analyzed in this study \cite{iturritxa2014}.
In-situ measurements of defoliation of trees (serving as a proxy for tree health) were collected by visual inspection of experts in the field to serve as the response variable \textit{defoliation} which ranges from 0 - 100 (in \%) (\autoref{fig:defol-distr}).

@@

\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.7\textwidth] {defoliation-distribution-plot-1.pdf}
		\caption{Response variable "defoliation at trees" for plots Laukiz1, Laukiz2, Luiando and Oiartzun. \texttt{n} corresponds to the total number of trees in the plot, $\bar{x}$ refers to the mean defoliation, respectively. Values for Laukiz1, Luiando and Oiartzun were observed in 5\% intervals; for Laukiz2 defoliation was observed at multiple heights and then averaged, leading to smaller defoliation differences than 5\%.}\label{fig:defol-distr}
	\end{center}
\end{figure}

@@

The \textit{Pinus radiata} plots of this study, namely Laukiz1, Laukiz2, Luiando and Oiartzun, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
Oiartzun consists of most tree observations (n = 559) while Laukiz2 shows the largest area size (1.44 ha).
All plots besides Luiando are located within 100 km from the coast (\autoref{fig:study_area}).
In total 1808 observations are available (Laukiz1 = 559, Laukiz2 = 451, Luiando = 301, Oiartzun = 497).
Field surveys were conducted in September 2016 by experienced forest pathologists.
Defoliation was measured via visual inspection using 5\% intervals with the help of a dedicated score card.
For Laukiz2, values at three height levels (bottom, mid, top) were available and averaged into an overall defoliation value, leading to values outside of the 5\% interval of the other three plots (e.g. 8.33 \%).
Estimating the human observer error of such surveys when assessing defoliation is an issue which is being discussed since many years \cite{innes1993}.
Even though no estimation error was recorded in this study, \cite{maclean1982} estimated human observer errors when assessing defoliation to range between 7\% - 18\%.

@@

The airborne hyperspectral data was acquired during two flight campaigns which took place at noon on September 28th and October 5th 2016.
Images were taken by an AISA EAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands were corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_metadata}.

@@

\begin{table}[t]
	\centering
	\caption[t]{Specifications of hyperspectral data.}
	\begingroup
	\begin{tabular}{ll}
		\\
		Characteristic         & Value                               \\
		\toprule
		Geometric resolution   & 1 m                                 \\
		Radiometric resolution & 12 bit                              \\
		Spectral resolution    & 126 bands (404.08 nm --- 996.31 nm) \\
		Correction:            & Radiometric, geometric, atmospheric
	\end{tabular}
	\endgroup\label{tab:hyperparameter_metadata}
\end{table}

@@

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated.
NRIs follow the \ac{OMNBR} concept of data-driven information extraction from narrow-band indices of hyperspectral data \cite{thenkabail2000,thenkabail2018}.
While various index formulations such as band ratios or normalized ratios are available, indices involving the same bands are strongly correlated.
Since the widely-used NDVI index belongs to the family of normalized ratio indices (NRIs), which are implemented in the \texttt{hsdar} R package, we used the following normalized difference index (NDI) formula to combine all pairs of reflectances:

@@

To reduce the feature space of a dataset, conceptually differing approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or \ac{PCA} \cite{bommert2020, das2001, guyon2003, jolliffe2016}.
In contrast to wrapper methods, filters can be added to the hyperparameter optimization step and have a lower computational footprint.
In addition, filters are less known than wrapper methods and in recent years ensemble filters were introduced which have shown promising results compared to single filter algorithms \cite{drotar2017a}.
These two points mainly led to the decision to focus on filter methods for this work and evaluate their effectiveness on highly-correlated, high-dimensional datasets.
Due to this focus, only this sub-group of feature selection methods will be introduced in greater detail in the following sections.

@@

% Filter methods
The concept of filters originates from the idea of ranking features following a score calculated by an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (e.g. numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase during model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a additional hyperparameter of the model.
The idea is to optimize the number of ranked features to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' can possibly enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in ML which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was used \cite{drotar2017a}.
Its final feature order is determined by the sum of all single filters ranks.

@@

\begin{itemize}
	\item Univariate/multivariate (scoring based on a single variable / multiple variables).
	\item Linear/non-linear (usage of linear/non-linear calculations).
	\item Entropy/correlation (scoring based on derivations of entropy or correlation-based approaches).
\end{itemize}

@@

\begin{itemize}
	\item  Extreme Gradient Boosting (XGBoost)
	\item  Random Forest (RF)
	\item  Penalized Regression (with L1 / lasso and L2 / ridge penalties)
	\item  Support Vector Machine (SVM, radial basis function Kernel)
	\item  Featureless learner
\end{itemize}

RF and SVM are well-established algorithms and widely used in (environmental) remote sensing.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmarking studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge regression does not remove variables from the model (penalization to zero) but it just shrinks them towards zero, keeping them in the model.
A featureless learner was included for a baseline comparison.

In total, the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.
The selected hyperparameter settings are shown in appendix \autoref{tab:hyperparameter_limits}.
All code and data are included in the research compendium of this study (\url{https://doi.org/10.5281/zenodo.2635403}).

@@

The idea of splitting the features into different sets originated from the question whether feature-engineered indices derived from reflectance values have a positive effect on model performance.
Peña et al. 2017 \cite{pena2017} is an exemplary study which used this approach in a spectro-temporal setting.
Benchmarking learners on these feature sets while keeping all other variables such as model type, tuning strategy and partitioning method constant makes it possible to draw conclusions on their individual impact.
Each feature set has distinct capabilities that differentiate it from the others.
This can have both a positive and negative effect on the resulting performance, which is one item this study aims to explore.
For example, feature set VI misses certain parts of the spectral range as the chosen indices only use specific spectral bands.
Feature set NRI will introduce highly-correlated features, for which some algorithms may be more suitable than others.
In addition to these feature sets, the following combinations of feature sets were also compared:

@@

Some individual features were removed before using the datasets for modelling when being numerically equivalent to another feature based on the pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates for feature set VI to 86 (from 89).
This decision was made to prevent issues for the subsequent tuning, filtering and model fitting steps which might occur when passing features with a pairwise correlation of (almost) one.
Additional variable selection was solely devoted to the respective feature selection methods in this study. <!-- FIXME: I don't understand this sentence. -->

@@

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage \cite{binder2020}.
For \ac{PCA}, the number of principal components was tuned.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$, a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e. $m_\textrm{try} = 1$) and 0.5 (i.e. $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features that were left after optimizing the feature percentage during tuning.

@@

In this work permutation-based feature importance was calculated to estimate feature importance / effects \cite{apley2019}.
With the limitations in mind when applied to correlated features, the aim was to get a general overview of the feature importance of the hyperspectral bands while trying to avoid an over-interpretation of results.
The best-performing algorithm on the HR task (i.e. SVM) was used for the feature importance calculation.

@@

For environmental interpretation purposes the ten most important indices of the best performing models of feature sets HR and VI were linked to the spectral regions of the hyperspectral data.
The aim was to visualize the most important features along the spectral curve of the plots to better understand which spectral regions were most important for the model.

@@

All tasks of this study were conducted using the open-source statistical programming language R \cite{rcoreteam2019}.
A complete list of all R packages used in this study can be found in the linked repositories mentioned in the next paragraph.
Due to space limitations only the selected packages with high impact on this work will be explicitly cited.

@@

% learner
% - SVM Car is the overall winner
% - Lasso quite stable (varies within 1 - 4), ridge a bit more variable (2 - 6)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Relief} filter achieved the best overall performance (RMSE of 28.09 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalties showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Differences to other feature sets were very small for lasso (below one p.p.) and a bit larger for ridge regression (between two and five p.p.).
XGBoost showed very poor performances for some feature sets and fills the ten last places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
The use of filters with RF resulted in a substantial increase in performance in all tasks with the exception of NRI, in which the difference among all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost's performance depended strongly on feature selection: In two out of six tasks, using no filter resulted in the worst or second worst performance.
XGBoost showed the highest overall differences between filters for a single task: for feature set HR, the range is up to 15 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

@@

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter achieved in only one instance (RF on HR-NRI-VI) the best performance among the used filters (\autoref{fig:filter-effects-borda}).
For RF and XGBoost it most often ranked within the better half among all filters of a specific task.
For XGBoost on the NRI task, the Borda filter scored the second worst performance.

% Selected percentage of features across plots
The number of features selected based on model tuning strongly varied across learners and plots.
While for RF, the least features were selected when Luiando or Oiartzun was the test set ($n = 1$), more than 90\% of all features were used for test sets Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set, (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots and among all learners, the single highest fraction being 24\% for test set Laukiz2.
In general, all learners behaved rather differently on the various plots, and therefore no overall pattern could be identified.

@@

The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For the HR dataset, most (i.e., six out of ten) relevant features clustered around the infrared region (920 nm - 1000 nm), while for VI, eight out of ten concentrate within the wavelength range of 700 nm - 750 nm (the so called \enquote{red edge}).
For HR, four features in the infrared region (920 nm - 1000 nm) were identified by the model to be most important, being assciated with a mean decrease in RMSE of around 1 p.p.
Overall, apart from the top five features, the vast majority of features showed only a small importance with average decreases in RMSE below 0.5 p.p.

% permutation based var imp for datasets HR and VI
\begin{figure*} [ht!]
	\centering
	\begin{center}
		\includegraphics[width=0.9\textwidth] {fi-permut-vi-hr-1.pdf}
		\caption{Variable importance for feature sets HR and VI: Mean decrease in RMSE for one-hundred feature permutations using the SVM learner. The wavelength range on the x-axis matches the range of the hyperspectral sensor (400 nm - 1000 nm). For each dataset, the ten most important features were highlighted as black dots and labeled by name. Grey dots represent features from importance rank 11 to last. The spectral signature (mean) of each plot was added as a reference on a normalized reflectance scale [0, 1] (secondary y-axis). VI features were decomposed into their individual formula parts; all instances being connected via dashed lines. Each VI feature is composed out of at least two instances.}\label{fig:fi-permut-vi-hr}
	\end{center}
\end{figure*}

@@

The best aggregated performance of this study (SVM + \enquote{Relief} filter, RMSE 28.12 p.p.) has to be seen in the light of model overfitting (see \autoref{subsec:perf-plot-char}).
Leaving out the performance on Laukiz2 when aggregating results, the mean RMSE would be around 19 p.p.
However, leaving out a single plot would also change the prediction results for the other plots because the observations from Laukiz2 would not be available for model training.
Due to the apparent presence of model overfitting in this study it can be postulated that more training data representing a greater variety of situations is needed.
A model can only make robust predictions if it has learned relationships across the whole range of the response.
Hence, care should be taken when predicting to the landscape scale using models fitted on this dataset due to their lack of generalizability caused by the limitations of the available training data.
However, when inspecting the fold level performances, it can be concluded that the models performed reasonably well predicting defoliation greater than 50\% but failed for lower levels.
This applied to all learners of this study.
The overall performance of all learners achieved in this study can be classified as "poor" given that only the SVM learner was able to outperform the featureless learner (\autoref{tab:best-learner-perf}).

In addition, data quality issues (\autoref{subsec:data-quality}) might have an influence on model performances.
These include the timing of the acquisition of the hyperspectral data (late phenological phase), field measurement errors when surveying defoliation, the influence of background reflectance (e.g. soil reflectance) and the possible positional offset of measured GPS coordinates of trees.

@@

One objective of this study was whether expert-based or data-driven feature engineering has a positive influence on model performance.
With respect to \autoref{fig:perf-result}, no overall positive or negative trend was found for all models that related to specific feature sets.
The performance of RF and XGBoost on the VI feature set was around four to six p.p. lower than on others.
One reason could be the lack of coverage in the wavelength area between 810 nm and 1000 nm (\autoref{fig:fi-permut-vi-hr}).
In addition, for all learners but SVM a better performance was observed when NRI indices were included in the feature set (i.e. NRI-VI, HR-NRI, HR-NRI-VI).

@@

Arbitrary seeming values of selected feature percentages like 99.97268 \% in \autoref{tab:tune-perc-sel-features} emerge from the specific setting which got elected as the best setting during the tuning step.
The specific value to benchmark during tuning is defined by the internal surrogate learner of the MBO tuning method in this study and can be any numeric value between 0 and 1 (interpreted as percentage of available features).

@@

Not surprisingly the most important features for both HR and VI datasets were identified around the red edge of the spectra, specifically in the range of 680 nm to 750 nm.

>>>>

% Explain what is new in this study and link to methodology
In recent years, various studies which have utilized hyperspectral data to analyze pest/fungi infections on trees were published:
Pinto et al. \citep{pinto2020} successfully used hyperspectral imagery to characterize pest infections on peanut leaves using random forest while Yu et. al \cite{yu2021} aimed to detect Pine wilt disease in pine plots in China using vegetation indices derived from hyperspectral data.
Other studies which applied hyperspectral data for forest health monitoring are \cite{lin2014,kayet2019,dash2017}.
Building upon these successful applications of hyperspectral remote sensing usage in the field of leaf and tree health monitoring, this work analyses tree defoliation in northern Spain using airborne hyperspectral data.
The methodology of this study uses ML methods in combination with feature selection and hyperparameter tuning.
In addition, feature importance was analyzed.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high dimensionality in environmental modeling \cite{trunk1979, xu2016}.

@@

Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata D. Don}) plantations in northern Spain.
The trees in the plots suffer from infections of pathogens such as \textit{Diplodia sapinea (Fr.) Fuckel}, \textit{Fusarium circinatum Nirenberg \& O'Donnell}, \textit{Armillaria mellea (Vahl) P. Kumm}, \textit{Heterobasidion annosum (Fr.) Bref}, \textit{Lecanosticta acicola (Thüm) Syd.} and \textit{Dothisthroma septosporum (Dorogin) M. Morelet} causing (among others) needle blight, pitch canker and root diseases \cite{mesanza2016, iturritxa2017}.
The first two fungi are mainly responsible for the foliation loss of the trees analyzed in this study \cite{iturritxa2014}.
In-situ measurements of defoliation of trees (serving as a proxy for tree health) were collected by visual inspection of experts in the field to serve as the response variable \textit{defoliation} which ranges from 0--100 (in \%) (\autoref{fig:defol-distr}).

@@

\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.7\textwidth] {defoliation-distribution-plot-1.pdf}
		\caption{Response variable ``defoliation at trees'' for plots~Laukiz1, Laukiz2, Luiando and Oiartzun. \texttt{n} corresponds to the total number of trees in the plot, $\bar{x}$ refers to the mean defoliation, respectively. Values for Laukiz1, Luiando and Oiartzun were observed in 5\% intervals; for Laukiz2 defoliation was observed at multiple heights and then averaged, leading to smaller defoliation differences than 5\%.}\label{fig:defol-distr}
	\end{center}
\end{figure}

@@

The \textit{Pinus radiata} plots of this study, namely Laukiz1, Laukiz2, Luiando and Oiartzun, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
Oiartzun consists of most tree observations (n = 559) while Laukiz2 shows the largest area size (1.44~ha).
All plots besides Luiando are located within 100~km from the coast (\autoref{fig:study_area}).
A total of 1808 observations are available (Laukiz1 = 559, Laukiz2 = 451, Luiando = 301, Oiartzun = 497).
Field surveys were conducted in September 2016 by experienced forest pathologists.
Defoliation was measured via visual inspection using 5\% intervals with the help of a dedicated score card.
For Laukiz2, values at three height levels (bottom, mid, top) were available and averaged into an overall defoliation value, leading to values outside of the 5\% intervals of the other three plots (e.g., 8.33\%).
Estimating the human observer error of such surveys when assessing defoliation is an issue which is being discussed since many years \cite{innes1993}.
Even though no estimation error was recorded in this study, \cite{maclean1982} estimated human observer errors when assessing defoliation to range between 7\% - 18\%.

@@

The airborne hyperspectral data was acquired during two flight campaigns which took place at noon on September 28th and October 5, 2016.
Images were taken by an AISA EAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands were corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_metadata}.

@@

\begin{table}[t]
	\centering
	\caption[t]{Specifications of hyperspectral data.}
	\begingroup
	\begin{tabular}{ll}
		\\
		Characteristic         & Value                               \\
		\toprule
		Geometric resolution   & 1~m                                 \\
		Radiometric resolution & 12 bit                              \\
		Spectral resolution    & 126 bands (404.08~nm~--- 996.31~nm) \\
		Correction:            & Radiometric, geometric, atmospheric
	\end{tabular}
	\endgroup\label{tab:hyperparameter_metadata}
\end{table}

@@

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated.
NRIs follow the \ac{OMNBR} concept of data-driven information extraction from narrow-band indices of hyperspectral data \cite{thenkabail2000,thenkabail2018}.
While various index formulations such as band ratios or normalized ratios are available, indices involving the same bands are strongly correlated.
Since the widely\ used NDVI index belongs to the family of normalized ratio indices (NRIs), which are implemented in the \texttt{hsdar} R package, we used the following normalized difference index (NDI) formula to combine all pairs of reflectances:

@@

To reduce the feature space of a dataset, conceptually differing approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or \ac{PCA} \cite{bommert2020, das2001, guyon2003, jolliffe2016}.
In contrast to wrapper methods, filters can be added to the hyperparameter optimization step and have a lower computational footprint.
In addition, filters are less known than wrapper methods and in recent years ensemble filters were introduced which have shown promising results compared to single filter algorithms \cite{drotar2017}.
These two points mainly led to the decision to focus on filter methods for this work and evaluate their effectiveness on highly\ correlated, high-dimensional datasets.
Due to this focus, only this subgroup of feature selection methods will be introduced in greater detail in the following sections.

@@

% Filter methods
The concept of filters originates from the idea of ranking features following a score calculated by an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (e.g., numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase during model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as an additional hyperparameter of the model.
The idea is to optimize the number of ranked features to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' can possibly enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in ML which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was used \cite{drotar2017}.
Its final feature order is determined by the sum of all single filters ranks.

@@

\begin{itemize}
	\item Univariate/multivariate (scoring based on a single variable/multiple variables).
	\item Linear/non-linear (usage of linear/non-linear calculations).
	\item Entropy/correlation (scoring based on derivations of entropy or correlation-based approaches).
\end{itemize}

@@

\begin{itemize}
	\item  Extreme Gradient Boosting (XGBoost)
	\item  Random Forest (RF)
	\item  Penalized Regression (with L1/lasso and L2/ridge penalties)
	\item  Support Vector Machine (SVM, radial basis function Kernel)
	\item  Featureless learner
\end{itemize}

RF and SVM are well-established algorithms and widely used in (environmental) remote sensing.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmarking studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly\ correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge regression does not remove variables from the model (penalization to zero) but it just shrinks them towards zero, keeping them in the model.
A featureless learner was included for a baseline comparison.

In total, the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3~ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.
The selected hyperparameter settings are shown in appendix \autoref{tab:hyperparameter_limits}.
All code and data are included in the research compendium of this study (\url{https://doi.org/10.5281/zenodo.2635403}).

@@

The idea of splitting the features into different sets originated from the question whether feature-engineered indices derived from reflectance values have a positive effect on model performance.
Peña et al. 2017 \cite{pena2017} is an exemplary study which used this approach in a spectro-temporal setting.
Benchmarking learners on these feature sets while keeping all other variables such as model type, tuning strategy and partitioning method steady makes it possible to draw conclusions on their individual impact.
Each feature set has distinct capabilities that differentiate it from the others.
This can have both a positive and negative effect on the resulting performance, which is one item this study aims to explore.
For example, feature set VI misses certain parts of the spectral range as the chosen indices only use specific spectral bands.
Feature set NRI will introduce highly\ correlated features, for which some algorithms may be more suitable than others.
In addition to these feature sets, the following combinations of feature sets were also compared:

@@

Some individual features were removed before using the datasets for modeling when being numerically equivalent to another feature based on the pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates for feature set~VI to 86 (from 89).
This decision was made to prevent issues for the subsequent tuning, filtering and model fitting steps which might occur when passing features with a pairwise correlation of (almost) one.
Additional variable selection was solely devoted to the respective feature selection methods in this study. <!-- FIXME: I don't understand this sentence. -->

@@

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage \cite{binder2020}.
For \ac{PCA}, the number of principal components was tuned.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$, a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e., $m_\textrm{try} = 1$) and 0.5 (i.e., $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features that were left after optimizing the feature percentage during tuning.

@@

In this work permutation-based feature importance was calculated to estimate feature importance/effects \cite{apley2019}.
With the limitations in mind when applied to correlated features, the aim was to get a general overview of the feature importance of the hyperspectral bands while trying to avoid an over-interpretation of results.
The best-performing algorithm on the HR task (i.e., SVM) was used for the feature importance calculation.

@@

For environmental interpretation purposes, the ten most important indices of the best performing models of feature sets HR and VI were linked to the spectral regions of the hyperspectral data.
The aim was to visualize the most important features along the spectral curve of the plots to better understand which spectral regions were most important for the model.

@@

All tasks of this study were conducted using the open-source statistical programming language R \cite{rcoreteam2019}.
A complete list of all R packages used in this study can be found in the linked repositories mentioned in the next paragraph.
Due to space limitations, only the selected packages with high impact on this work will be explicitly cited.

@@

% learner
% - SVM Car is the overall winner
% - Lasso quite stable (varies within 1 - 4), ridge a bit more variable (2 - 6)
% - all MBO penalized methods have no difference across tasks
SVM combined with the \enquote{Relief} filter achieved the best overall performance (RMSE of 28.09 p.p.) (\autoref{tab:perf-top-10}).
%L1 penalized regression showed slightly better results in three out of six tasks when being optimized using the package internal 10-fold CV compared to MBO (RMSE of 50 vs 58).
Regression with ridge (L2) and lasso (L1) penalties showed their best performances on the NRI feature set (\autoref{tab:best-learner-perf}).
Differences to other feature sets were very small for lasso (below one p.p.) and a bit larger for ridge regression (between two and five p.p.).
XGBoost showed very poor performances for some feature sets and fills the last\ ten places of the ranking (\autoref{tab:perf-worst-10}).
Especially when combined with PCA, the algorithm was not able to achieve adequate performances.

% filters
% - influence varies based on algorithm and task (SVM no influence, RF medium, XGB medium)
% - no substantial decrease in performance when using filters
% - no advantage of ensemble borda filter compared to others
Effects of filter methods on performance differed greatly between algorithms:
SVM showed no variation in performance across filters (\autoref{fig:filter-effects-no-filter}).
The use of filters with RF resulted in a substantial increase in performance in all tasks with the exception of NRI, in which the difference between all filters was also the smallest (\autoref{fig:filter-effects-no-filter}).
XGBoost's performance depended strongly on feature selection: In two out of six tasks, using no filter resulted in the worst or second\-/worst performance.
XGBoost showed the highest overall differences between filters for a single task: for feature set HR, the range is up to 15 p.p. (\enquote{CMIM} vs. \enquote{no filter}) (\autoref{fig:filter-effects-no-filter}).

@@

% Borda vs. other filters
% - most often among the first best filters for each task
% - never the best
The Borda filter achieved in only one instance (RF on HR-NRI-VI) the best performance of\ all filters (\autoref{fig:filter-effects-borda}).
For RF and XGBoost, it most often ranked within the better half among all filters of a specific task.
For XGBoost on the NRI task, the Borda filter scored the second\-/worst performance.

% Selected percentage of features across plots
The number of features selected based on model tuning strongly varied across learners and plots.
While for RF, the least features were selected when Luiando or Oiartzun was the test set ($n = 1$), more than 90\% of all features were used for test sets~Laukiz1 and Laukiz2 (\autoref{tab:tune-perc-sel-features}).
RF and SVM used very few features only when Laukiz1 was the test set (three and one, respectively) whereas this was the plot for which XGBoost used almost all features ($n = 1226$).
Overall, SVM used the least features across all plots and among all learners, the highest\ single fraction being 24\% for test set~Laukiz2.
In general, all learners behaved rather differently on the various plots, and therefore no overall pattern could be identified.

@@

The most important features for datasets HR and VI showed an average decrease in RMSE of 1.06 p.p. (HR, B69) and 1.93 p.p. (VI, Vogelmann2) (\autoref{fig:fi-permut-vi-hr}).
For the HR dataset, most (i.e., six out of ten) relevant features clustered around the infrared region (920~nm - 1000~nm), while for VI, eight out of ten concentrate within the wavelength range of 700~nm - 750~nm (the so\-/called \enquote{red edge}).
For HR, four features in the infrared region (920~nm - 1000~nm) were identified by the model to be most important, being associated with a mean decrease in RMSE of around 1 p.p.
Overall, apart from the top five features, the vast majority of features showed only a small importance with average decreases in RMSE below 0.5 p.p.

% permutation based var imp for datasets HR and VI
\begin{figure*} [ht!]
	\centering
	\begin{center}
		\includegraphics[width=0.9\textwidth] {fi-permut-vi-hr-1.pdf}
		\caption{Variable importance for feature sets HR and VI: Mean decrease in RMSE for one\ hundred feature permutations using the SVM learner. The wavelength range on the x-axis matches the range of the hyperspectral sensor (400~nm - 1000~nm). For each dataset, the ten most important features were highlighted as black dots and labeled by name. Gray dots represent features from importance rank~11 to last. The spectral signature (mean) of each plot was added as a reference on a normalized reflectance scale [0, 1] (secondary y-axis). VI features were decomposed into their individual formula parts; all instances being connected via dashed lines. Each VI feature is composed out of at least two instances.}\label{fig:fi-permut-vi-hr}
	\end{center}
\end{figure*}

@@

The best aggregated performance of this study (SVM + \enquote{Relief} filter, RMSE~28.12 p.p.) has to be seen in the light of model overfitting (see \autoref{subsec:perf-plot-char}).
Leaving out the performance on Laukiz2 when aggregating results, the mean RMSE would be around 19 p.p.
However, leaving out a single plot would also change the prediction results for the other plots because the observations from Laukiz2 would not be available for model training.
Due to the apparent presence of model overfitting in this study, it can be postulated that more training data representing a greater variety of situations is needed.
A model can only make robust predictions if it has learned relationships across the whole range of the response.
Hence, care should be taken when predicting to the landscape scale using models fitted on this dataset due to their lack of generalizability caused by the limitations of the available training data.
However, when inspecting the fold\-/level performances, it can be concluded that the models performed reasonably well predicting defoliation greater than 50\% but failed for lower levels.
This applied to all learners of this study.
The overall performance of all learners achieved in this study can be classified as ``poor'' given that only the SVM learner was able to outperform the featureless learner (\autoref{tab:best-learner-perf}).

In addition, data quality issues (\autoref{subsec:data-quality}) might have an influence on model performances.
These include the timing of the acquisition of the hyperspectral data (late phenological phase), field measurement errors when surveying defoliation, the influence of background reflectance (e.g., soil reflectance) and the possible positional offset of measured GPS coordinates of trees.

@@

One objective of this study was whether expert-based or data-driven feature engineering has a positive influence on model performance.
With respect to \autoref{fig:perf-result}, no overall positive or negative trend was found for all models that related to specific feature sets.
The performance of RF and XGBoost on the VI feature set was around four to six p.p. lower than on others.
One reason could be the lack of coverage in the wavelength area between 810~nm and 1000~nm (\autoref{fig:fi-permut-vi-hr}).
In addition, for all learners but SVM a better performance was observed when NRI indices were included in the feature set (i.e., NRI-VI, HR-NRI, HR-NRI-VI).

@@

Arbitrary seeming values of selected feature percentages like 99.97268\% in \autoref{tab:tune-perc-sel-features} emerge from the specific setting which got elected as the best setting during the tuning step.
The specific value to benchmark during tuning is defined by the internal surrogate learner of the MBO tuning method in this study and can be any numeric value between 0 and 1 (interpreted as a\ percentage of available features).

@@

Not surprisingly the most important features for both HR and VI datasets were identified around the red edge of the spectra, specifically in the range of 680~nm to 750~nm.
